---
title: "Predicting the Manner of Exercise"
author: "Jim VanOosten"
date: "December 22, 2015"
output: html_document
---

## Summary

The goal of this project is to use exogenous data to predict the manner in which an exercise was performed.  Training and testing data sets were provided for the project.  After loading the training data, it is tidied by removing the columns that are not used in the model training.  The training data was split for cross validation into a training set (75%) and a probe set (25%).  The model was trained using the Random Forest algorithm and its accuracy was found to be over 99% using the probe data.  As a final verification, the model was used to correctly predict 20 tests provided in the testing data set. 

## Data Processing

This section includes downloading, tidying, and splitting the data for cross validation.   The data for this project came from a research project focused on Human Activity Recognition (HAR).  The participants in the study were asked to perform bicep curls correctly and incorrectly in 5 different ways.  Each way is considered a class, A-F, and recorded in the classe variable of the data set.

* Class A: exactly according to the specification
* Class B: throwing the elbows to the front
* Class C: lifting the dumbbell only halfway
* Class D: lowering the dumbbell only halfway 
* Class E: throwing the hips to the front 

The sensor data was recorded by accelerometers on the participant's belt, forearm, arm, and dumbell.  For more detail see  http://groupware.les.inf.puc-rio.br/har.  

### Loading the data

The training data is downloaded using the provided URL and put into a dataframe object.

```{r loadtraindata, echo = TRUE}
  setwd("~/Documents/coursera/datascience/predmachlearn/project/practicalmachinelearning")
  set.seed(327)

  # Verify the pml-training.csv has been downloaded 
  if (!file.exists("pml-training.csv")) {
    dataUrl = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    download.file(dataUrl, destfile = "pml-training.csv", method = "curl")
  }

  # load the data in the pmltrain dataframe
  pmltrain <- read.csv("pml-training.csv", header = TRUE)
  dim(pmltrain)
```

### Predictor Reduction

The training data is tidied by removing columns not used to train the model.  The discussion forums (https://class.coursera.org/predmachlearn-035/forum/thread?thread_id=2) revealed that the test data columns that had NA values could not be used as predictors as they would cause errors if included in the model.  So those columns are pruned from the training data.  

```{r tidytraindata, echo = TRUE}
  library(dplyr)
  # Verify the pml-testing.csv has been downloaded 
  if (!file.exists("pml-testing.csv")) {
    dataUrl = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    download.file(dataUrl, destfile = "pml-testing.csv", method = "curl")
  }

  # load the data in the test dataframe
  pmltest <- read.csv("pml-testing.csv", header = TRUE)
  dim(pmltest)
  
  # get a list of the test columns with NA in the column
  na_names <- names(pmltest[, colSums(is.na(pmltest)) == nrow(pmltest)])
  
  # remove columns from the training set that are NA in the test set 
  pmltrain <- pmltrain[,-which(names(pmltrain) %in% na_names)]
```

In addition, I removed user_name, time stamps, and window columns as predictors so the model would be based solely on the sensor data.  

```{r reducepredicters, echo = TRUE}
  library(dplyr)
  # only keep the sensor based columns and the classe 
  pmltrain <- select(pmltrain, roll_belt:classe)
```

### Data Exploration

Exploring the sensor data provides clues on how it can be used to classify the exercises.  Here I plot the roll_belt vs pitch_belt while using the classe to color the data.    

```{r exploredata, echo = TRUE}
  library(ggplot2)
  qplot(roll_belt,pitch_belt,colour=classe,data=pmltrain)
```

Class E is distinguished from the other classes in the plot which makes sense since the accelerometer on the belt would record the action of throwing the hips to the front.

### Cross Validation 

The training set is split into two data sets. One for training and one for probing the model before the final test. 

```{r splitdata, echo = TRUE}
  library(caret)
  inTrain <- createDataPartition(y=pmltrain$classe, p=0.75, list=FALSE)
  training <- pmltrain[inTrain,] 
  dim(training)
  probing <- pmltrain[-inTrain,] 
  dim(probing)
```

## Model Creation and Verification 

### Model Creation 

I selected Random Forest algorithm to create the model.  The Random Forest algorithm grows many classification trees, forming a forest, from which the tree with the most votes (based on its classification results) is selected.  All of the sensor based columns that remain in the training data set are used as predictors while the classe variable is the outcome.

The doMC package is used to exploit the multiple cores on my work station during the model creation.  The doMC uses the parallel package to fork processes and split computations.  As a result, the model creation completes in a fraction of the time versus the single core execution.

```{r trainmodel, echo = TRUE, cache=TRUE}
  library(doMC)
  registerDoMC(cores = 6)
  modFit <- train(classe ~ ., data=training, method="rf")
```

A plot of the model shows most of the accuracy is achieved with a few predictors. As a follow on excercise, I could explore reducing the predictors using pre-processing techniques like Principle Component Analysis (PCA). 

```{r plotmodel, echo = TRUE}
  plot(modFit)
```


### Model Verification

The probe data set aside for cross validation is used to gauge the model accuracy.  The confusion matrix shows the matrix of the predictions versus the reference Classes A-E.  The diagonal represents correct predictions and the rows show which category a given prediction actually belonged in.  The accuracy of the model was over 99%.  Statistics are provided on a per Class basis. 

```{r verifymodel, echo=FALSE}
  predictions <- predict(modFit, probing)
  confusionMatrix(predictions,probing$classe)
```

## Testing 

The machine learning algorithm created above is applied to the 20 test cases provided in the test data.   The results are 100% correct.  The results are saved to files for project submission.

```{r, finalpredictions, echo=FALSE}
  predictfinal <- predict(modFit, pmltest)
  predictfinal

  results <- "./results"
  if(!file.exists(results)) {dir.create(results)}

  pml_write_files = function(x,y){
    n = length(x)
    for(i in 1:n){
      filename = paste(y,"/problem_id_",i,".txt",sep="")
      write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
  }

  pml_write_files(predictfinal,results)  
```


